---
title: "Why Risk it, When You Can {rix} it: A Tutorial for Computational Reproducibility Focused on Simulation Studies"
shorttitle: "Reproducibility with {rix}"
author:
  - name: Felipe Fontana Vieira
    corresponding: true
    orcid: 0009-0006-0949-6569
    email: felipe.fontanavieira@ugent.be
    affiliations:
      - name: Ghent University 
        department: Department of Data Analysis 
    roles:
      - Conceptualization
      - Methodology
      - Software
      - Data curation
      - Formal analysis
      - Visualization
      - Writing - original draft
      - Writing - review & editing
  - name: Jason Geller
    orcid: 0000-0002-7459-4505
    affiliations:
      - name: Boston College
        department: Department of Psychology and Neuroscience
    roles: 
      - Conceptualization
      - Supervision
      - Validation 
      - Writing - review & editing
  - name: Bruno Rodrigues
    orcid: 0000-0002-3211-3689
    affiliations:
      - name: Ministry of Research and Higher Education, Luxembourg
        department: Statistics and Data Strategy Departments
    roles: 
      - Conceptualization
      - Software
      - Writing - review & editing
      
abstract: |
  Computational reproducibility remains limited in psychological research,
  despite widespread norms for sharing data and analysis code. These materials
  alone rarely suffice: re-executing an analysis requires reconstructing the
  computational environment it depends on, from the programming language version
  and packages to system-level dependencies. In this article, we introduce Nix 
  and the {rix} R package as a practical framework for achieving computational 
  reproducibility in simulation-based research. We provide a 
  step-by-step tutorial demonstrating how {rix} can be
  used to define, build, and share isolated, project-specific software
  environments that precisely capture R versions, package dependencies, system
  libraries, and integrated development environments. We further illustrate
  this workflow by reproducing a complete manuscript using Quarto and the
  {apaquarto} extension, showing how analyses, figures, and text can be
  regenerated in a single, executable pipeline. Together, these tools lower the
  technical barrier to robust, end-to-end reproducibility and offer a scalable
  solution for simulation studies and methodological research in psychology and
  related fields.

keywords: [reproducibility, Nix, simulation studies, R, computational methods]
word-count: true
author-note:
  disclosures:
    gratitude: "We would like to thank Julia Rohrer, Ole Schacht and Dries 
    Debeer for their helpful feedback on earlier versions of this manuscript, 
    and Matteo Quartagno for feedback on seeds and reproducibility. 
    During the first round of revision, Claude Sonnet 4.6
    (Anthropic) was used to improve the grammar and coherence solely of the 
    Introduction and Discussion."
    conflict-of-interest: "Bruno Rodrigues is one of the authors and maintainers
     of the {rix} package presented in this tutorial. The authors have no other 
    conflicts of interest to declare."
    data-sharing: "Data and materials from this study can be accessed at 
    <https://doi.org/10.5281/zenodo.21098114>."
format:
  apaquarto-pdf:
    document-mode: man
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{setspace}
        \usepackage{etoolbox}
        \raggedbottom
        % Single-space code blocks
        \AtBeginEnvironment{Shaded}{\singlespacing}
        \AtBeginEnvironment{verbatim}{\singlespacing}
  apaquarto-html:
    embed-resources: true
    toc: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 80
---

```{r}
#| label: load-packages
#| echo: false
#| message: false

# required packages
library(marginaleffects)
library(simhelpers)
library(rvinecopulib)
library(doParallel)
library(doRNG)
library(ggplot2)
library(cowplot)
library(knitr)
library(dplyr)
```

Psychological science is in the midst of a credibility revolution [@vazire2018].
However, to date, much of the attention has focused on empirical studies. Monte
Carlo simulation studies (hereafter referred to as simulation studies) have
received far less [@siepe_etall_2024], even though they are a primary tool for
evaluating the statistical methods [@morris_etall_2019] and may suffer from
selective reporting [@pawel_etall_2024; @bouma_etall_2026; @pawel_etall_2025].
One underexamined aspect for simulations is reproducibility (i.e.,
exactly reproducing a study's results from the provided data, code, materials,
and software) [@hardwicke2020]. In a simulation study, this involves more than
reanalyzing a fixed dataset, because the code also generates the data.
Reproducing such a study therefore involves the data generation process, the
analysis, and the summary of results. @montoya_anderson_2026 conceptually
address each of these stages.

Reproducibility in this sense is not all-or-nothing but occupies a continuum
[@peng_2011], on which a study is positioned according to the extent of the
materials that are publicly available. This ranges from publication alone at the
low end to code and data that are linked and directly executable at the high
end. Reaching that executable end is often treated as *full* reproducibility
[@peikert_brandmaier_2021; @peikert_etall_2021], and open science initiatives
have pushed work toward it [@kidwell_etall_2016; @levenstein_lyle_2018;
@nosek_etall_2015]. Yet code and data are never fully sufficient [@epskamp_2019;
@wiebels_moreau_2021; @ziemann_etall_2023; @brodeur_etall_2026]. Even error-free
code depends on a hierarchy of software components, including the programming
language version, the packages used in the analysis, and the system-level
dependencies those packages rely on. We define these and related terms in
@tbl-glossary. When they differ from the originals, code may fail, behave
inconsistently across machines, or yield conflicting numerical results
[@baker_etall_2024; @hodges_etall_2023; @glatard_etall_2015; @nosek_etall_2022].
Such problems are particularly acute for simulation studies, which rely on
interconnected code, package versions, and the broader software environment to
generate their data and results [@luijken_etall_2024; @siepe_etall_2024].

To make this concrete, we use the *computational environment* to refer to the
complete software context required for a study to run successfully. This entails
the programming language version, package versions, system-level dependencies,
and operating system (@fig-environment) [@rodrigues_2023;
@rodrigues_baumann_2026_polyglot]. We define *computational environment
reproducibility* as the ability to reconstruct these dependencies on top of
whatever operating system a machine runs, at any future time, such that
executing the same code yields the same numerical results[^1]. Current practice
falls short. Indeed, @siepe_etall_2024 report that, among the 100 simulation
studies they identified in psychology journals in 2021 and 2022, nearly
two-thirds provide no code, and those that do rarely document the computational
environment. This gap is consequential: because simulation studies inform
methodological recommendations, insufficient reproducibility may undermine
confidence in them [@luijken_etall_2024; @white_etall_2024].

[^1]: We note our characterization of computational environment reproducibility
    is not the most rigorous. For stricter definitions, we refer the interested
    reader to @malka2026docker (p. 3) and @lamb_zacchiroli_2022.

```{r}
#| label: fig-environment
#| fig-cap: "The computational environment as a set of nested layers,
#| each running on top of the layer below, based on Peikert & Brandmaier
#| (2021)."
#| echo: false
#| message: false
#| fig-width: 7
#| fig-height: 4

# outer nested layers: operating system and system-level dependencies
env_outer <- data.frame(
  xmin = c(0.0, 0.5),
  xmax = c(20.0, 19.5),
  ymin = c(0.0, 0.8),
  ymax = c(10.0, 8.6),
  label = c(
    "Operating system",
    "System-level dependencies (e.g., LAPACK)"
  ),
  fill = c("grey93", "grey86"),
  stringsAsFactors = FALSE
)
env_outer$labx <- (env_outer$xmin + env_outer$xmax) / 2

# two parallel components inside the system-level dependencies
env_inner <- data.frame(
  xmin = c(1.2, 10.4),
  xmax = c(9.6, 18.8),
  ymin = c(1.6, 1.6),
  ymax = c(7.8, 7.8),
  label = c(
    "Language version (e.g., R 4.3.3)",
    "Dynamic document tools (e.g., LaTeX)"
  ),
  fill = c("grey78", "grey78"),
  stringsAsFactors = FALSE
)
env_inner$labx <- (env_inner$xmin + env_inner$xmax) / 2

# R packages nested inside the language version
env_pkg <- data.frame(
  xmin = 1.7,
  xmax = 9.1,
  ymin = 2.2,
  ymax = 6.2,
  label = "R packages\n(e.g., ggplot2)",
  fill = "grey68",
  stringsAsFactors = FALSE
)
env_pkg$labx <- (env_pkg$xmin + env_pkg$xmax) / 2

ggplot() +
  geom_rect(
    data = env_outer,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill),
    color = "grey20",
    linewidth = 0.4
  ) +
  geom_text(
    data = env_outer,
    aes(x = labx, y = ymax - 0.4, label = label),
    fontface = "bold",
    size = 3.2
  ) +
  geom_rect(
    data = env_inner,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill),
    color = "grey20",
    linewidth = 0.4
  ) +
  geom_text(
    data = env_inner,
    aes(x = labx, y = ymax - 0.4, label = label),
    fontface = "bold",
    size = 3.0
  ) +
  geom_rect(
    data = env_pkg,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill),
    color = "grey20",
    linewidth = 0.4
  ) +
  geom_text(
    data = env_pkg,
    aes(x = labx, y = (ymin + ymax) / 2, label = label),
    fontface = "bold",
    size = 3.0
  ) +
  scale_fill_identity() +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL) +
  theme_void() +
  theme(plot.margin = margin(8, 8, 8, 8))
```

These challenges may persist for several reasons. Arguably, one is that even
researchers aware of these demands face a fragmented landscape of solutions,
each addressing only part of the problem. Package-level managers such as {renv}
[@ushey_2024] and {groundhog} [@simonsohn_2020] stabilize R package versions but
do not manage the R interpreter or the system-level dependencies those
packages depend on. Analytical pipeline tools such as {targets} [@landau_2021]
and Make [@feldman_1979] support reproducibility in a different sense: they
specify and automate the structure of an analysis by formalizing the order in
which steps run and by tracking dependencies among intermediate results.
These tools clarify *how* an analysis proceeds, but they assume that the
software stack required to run each step is already stable. Containerization
tools such as Docker, including R-focused implementations like Rocker
[@boettiger_2015; @boettiger_eddelbuettel_2017] that provide preconfigured
Docker images tailored to common R workflows, offer a more comprehensive
approach by bundling nearly the entire environment into a single executable
image. Yet their
use requires familiarity with Linux system administration, and even
containerization may suffer from temporal drift when Dockerfiles rely on mutable
upstream repositories [@malka2026docker]. For a detailed comparison of these
tools and their limitations, see @rodrigues_baumann_2026_polyglot. Researchers
thus face a trade-off between accessible but incomplete solutions and powerful
but technically demanding ones.

In this article, we focus specifically on computational environment
reproducibility as the foundation upon which other reproducibility practices
depend. For that, we introduce Nix [@dolstra_etall_2004], a software ecosystem
designed to make software installation deterministic, and {rix}
[@rodrigues_baumann_2025], an R interface that allows researchers to use Nix
without needing deep knowledge of its underlying language or infrastructure. Our
main objective with the tutorial is not to advocate for specific analytical
pipeline tool or to prescribe a particular workflow structure[^2]. Instead, we
aim to show what Nix and {rix} are and how they can establish a stable
environment within which simulation studies—whether organized in `.R`
files that `source()` others, through formal pipeline tools (e.g., {targets}) or
embedded as code chunks in `.Rmd` or `.qmd` files—can be executed reliably. We
illustrate these ideas through a reproducible simulation study conducted in R,
culminating in this automated APA-formatted manuscript generated with
`apaquarto` [@schneider_2024]. Although the example centers on R, the principles
underlying environment reproducibility with {rix} apply equally to other
languages, including Python and Julia, and to multiple development environments
such as RStudio, VS Code, or Positron.

[^2]: The analytical pipeline is one part of the broader reproducible workflow.
    We leave it aside for two reasons. First, many tools address it, each with
    its own conventions, and doing them justice would require a more tailored
    treatment. Second, although a formal analytical pipeline may be necessary
    for full reproducibility, it rests on a stable computational environment.
    Arguably, a study whose scripts are well documented, whose file dependencies
    are explained, and whose environment is stable therefore already sits high
    on the reproducibility spectrum, and that foundation is what we focus on
    here. For readers interested in the broader reproducible workflow, we
    suggest @peikert_etall_2021, @vanlissa_etall_2021, and @rodrigues_2023.

The remainder of this article is structured as follows. We first illustrate a
simulation study and the reproducibility risks associated with it. We then
introduce Nix and {rix}, along with their core principles. This is
followed by a step-by-step tutorial that (re)builds the simulation study as a
reproducible project with {rix}, from installing Nix to (re)generating this
manuscript. Last, we turn to additional considerations for reproducibility of 
simulation studies and close with a discussion.

| Term | Definition |
|:-----------------------------|:-------------------------------------------------|
| **System-level dependencies** | Software provided outside of R (by the operating system or a system package manager) that an analysis relies on, including shared libraries (e.g., LAPACK) and tools (e.g., Pandoc, LaTeX). |
| **Operating system** | The core software (e.g., Windows, macOS, Linux) that manages a computer's hardware and provides the platform on which all other programs, including R, run. |
| **R interpreter** | The R program that reads and executes your R code [@adler_2012]. Two machines may report the same R version but use interpreters compiled differently or linked against different system-level libraries [see @rcoreteam_installation, Section A.3]. |
| **Compiled** | Converted from human-readable source code into machine code by a compiler so the computer can run it directly. |
| **Analytical pipeline** | The sequence of analysis steps and the dependencies among them. Pipeline tools formalize this sequence, automate its execution, and re-run only the steps affected by a change. |
| **Software stack** | The full set of software layers an analysis depends on: the operating system, system-level dependencies, the R interpreter, and the R packages built on top. Note that this stack itself depends on the underlying hardware. |
| **Hardware** | The physical components of a computer (e.g., CPU, memory, and storage) that execute instructions and support the operation of software. |
| **Containerization** | Packaging software and its dependencies into an isolated, lightweight, shareable unit (a container), using tools such as Docker [@wiebels_moreau_2021]. |
| **Mutable upstream repositories** | External sources a build pulls from (e.g., base images or package repositories) that can change or disappear over time, meaning the same build instructions may later produce a different result or fail [@malka2026docker]. |
| **Reproducible workflow** | Overall research process that makes a study reproducible end to end, including version control, analytical pipeline, and a frozen computational environment. |
| **Literate programming** | Writing an analysis as a single document that combines narrative text and executable code (e.g., R Markdown or Quarto), keeping the report and its code together. |
| **Functional programming language** | A language designed so that the same inputs always produce the same output. |
| **Binary cache** | A server that stores prebuilt packages, so Nix can download a finished copy instead of compiling it from source. |
| **Cryptographic hash** | A fixed-length string of letters and digits computed from digital content, acting as its fingerprint. The same content always yields the same string, whereas any change to the content, however small, yields a different one. |

: Glossary of computing terms used in the main text. {#tbl-glossary}

# Setting the Stage: A Simulation Study and Its Reproducibility Risks

Imagine you have just been awarded a grant to conduct a large-scale simulation
study, where you evaluate the performance of a statistical estimator
under varying data-generating conditions (see Appendix A for full technical
details). This tutorial is organized around this example to ground our
discussion in a typical methods section, but readers can follow the tutorial
without engaging deeply with the simulation itself. In practice, simulation
studies are typically organized into multiple component files. To mimic that, 
our simulation is organized into five sequential scripts: data
generation (`01_data_generation.R`), model specification (`02_models.R`),
simulation execution (`03_run_simulation.R`), performance metric calculation
(`04_performance_metrics.R`), and results visualization (`05_plots.R`).
The `.R` scripts are available in the repository 
(<https://doi.org/10.5281/zenodo.21098114>; Simulation subfolder)[^3].

[^3]: For simplicity, we included all code as executable chunks in a single
    `.qmd` file. This entails that the simulation runs from start to finish
    (i.e., generating data, producing results and figures). This is impractical
    for many real-world simulation studies, which are often too computationally
    intensive. One possible solution, which we illustrate, is to run the 
    simulation outside the `.qmd` file and load the results for the analyses 
    in the same file.

Suppose a researcher attempts to reproduce the simulation run and results
reported in this article. What might prevent them from obtaining identical
outcomes, assuming error-free code? The natural first concern is the version of
packages. Installing R packages at a later time may lead to errors if functions
have been renamed or deprecated (e.g., `dplyr::cur_data()` changed to `pick()`,
or, although usually discouraged, internal functions, such as
`lavaan:::lav_utils_get_ancestors`, renamed to
`lavaan:::lav_graph_get_ancestors`), or to subtle differences in the results due
to changes in default settings (e.g., `stringsAsFactors` defaulting to FALSE as
of R 4.0). Beyond package versioning, many packages rely on system-level
dependencies that R does not provide. Our project illustrates this:
{rvinecopulib} [@rvinecopulib] requires a C++ compiler to install from source,
and {svglite} [@svglite] links against system libraries (e.g., libpng) that R
cannot install itself.

The R language version introduces another layer of dependency. Code written for
a recent version of R may rely on syntax or functionality that is unavailable in
earlier versions (e.g., the native pipe `|>` introduced in R 4.1). More subtly,
changes to R’s random number generation across versions mean that identical code
executed with the same seed can nevertheless produce different random sequences
[@ottoboni_stark_2018]. This is especially consequential for simulation studies,
because the random draws are not incidental to the analysis but constitute the
data itself. Therefore, the random number generator is part of the computational
environment. Moreover, generating data and reproducing every statistical
estimate *exactly* (down to the certain digits) can depend on lower-level
components still, such as the system-level libraries R uses for matrix
computations (e.g., BLAS/LAPACK). The potential solution we propose pins these
libraries, and increases the chances of reaching such a level of
reproducibility, consistent with empirical findings in other contexts
[@malka2026docker]. However, it must be noted that reproducing precisely every
number is hard to guarantee in every case, especially across different operating
systems and processors [e.g., @stan_reproducibility]. Although such differences 
are usually negligible,
appearing only in the last digits, the finite precision of floating-point
arithmetic [@goldberg1991; @ahrens2020] and the algorithm a specific R package
uses to generate data can occasionally amplify them into larger discrepancies.
With respect to all of this, we do not claim that this exact level of identity
must be met. How much reproducibility is required is ultimately for the
scientific community to decide (e.g., through journals and norms),
and reproducing the broader pattern of data generation and results is arguably 
enough (see also @montoya_anderson_2026, for a stage-by-stage approach to 
assessing reproducibility in simulation studies)[^4].

[^4]: The following illustration originated and it is credited to Danielle 
    Navarro. With the same seed and covariance matrix, `MASS::mvrnorm()` can 
    return different values on
    macOS than on Linux or Windows, whereas `mvtnorm::rmvnorm()` stays
    consistent. The difference traces to a sign flip in an eigenvector computed
    through the system's BLAS/LAPACK libraries, to which `mvtnorm` is invariant
    but `MASS` is not. The draws still come from the same distribution, so
    conclusions would generally be unchanged in simulation with multiple
    repetitions, but from a *pure* reproducibility standpoint this may be
    concerning. Pinning the numerical library with Nix (here, OpenBLAS) removes, 
    for this example, the discrepancy for `MASS::mvrnorm()`. For the full 
    demonstration, see the repository.

Finally, analyses embedded in a literate programming workflow add further
dependencies. Rendering R Markdown (`.Rmd`) or Quarto (`.qmd`) documents, for
example, requires a document conversion tool (such as Pandoc) and a typesetting
system (such as LaTeX or Typst), each with its own version and installation
requirements.

# Nix and {rix}: A Comprehensive Solution

A potential solution is Nix [@dolstra_etall_2004], a software ecosystem built so
that installing software gives the same result every time. Nix is a functional
programming language, a build tool, and a package manager wrapped in one. This
article focuses on the latter. Most package managers (think of Apple’s or
Android’s app stores) are imperative: they modify a system’s state as they
install or update software. Nix instead builds each environment from an explicit
description and never alters it afterward. Hence, the same description produces 
the same self-contained environment on other machines. Researchers state which
version of R and which packages they need, and Nix works out the underlying
system-level dependencies automatically.

## Core Principles

Rather than installing software into shared system folders (e.g., `/usr/lib`),
Nix gives every package its own folder under `/nix/store`. The folder's name
includes a short fingerprint (a *cryptographic hash*) computed from everything
that went into the package, its source code, dependencies, and build
instructions, so any change to those inputs yields a different folder. Because
each package is identified by its exact inputs this way, several versions of the
same software can sit side by side without conflict: a researcher can, for
example, keep R 4.1.0 and R 4.3.3 on the same machine, or use different package
versions across analyses [@rodrigues_baumann_2025].

The Nix ecosystem is built around `nixpkgs`, a version-controlled repository
comprising more than 120,000 packages, including nearly all of CRAN and
Bioconductor. By pinning a specific commit or date, researchers freeze the
entire software stack (i.e., R itself, R packages, and all system-level
dependencies). This eliminates the system-dependency problems that tools like
{renv} cannot address [@rodrigues_baumann_2025]. Empirical studies of historical
nixpkgs snapshots report that old environments can still be rebuilt years later
[@Malka_2024; @rodrigues_baumann_2026_polyglot]. Moreover, because prebuilt
binaries can be downloaded from a shared binary cache instead of compiled from
scratch, these environments are often ready in seconds, which keeps Nix
practical for everyday interactive work [@rodrigues_baumann_2025].

## The {rix} Package: R Interface to Nix

Because the Nix package manager is declarative, installing software requires
writing expressions in its language. Since this language is unfamiliar to most
researchers, we recommend and focus on using {rix} to lower this barrier. The
{rix} package provides an R-native interface: a single call to `rix()` generates
complete Nix configurations from standard R syntax, specifying R versions, CRAN
packages, system-level dependencies, and even Python or Julia components when
required. Users never need to read or write Nix code directly, as {rix}
translates automatically. For more on {rix}, including its integration with
rstats-on-nix (a curated fork of the Nix package collection whose dated
snapshots let users request the exact R and package versions available on a
chosen day), see @rodrigues_baumann_2025 and @rodrigues_baumann_2026_polyglot.

# A Practical Example: Setting up a Reproducible Simulation Study with {rix}

We take the perspective of a researcher (e.g., on macOS) beginning a simulation
study who wants it reproducible from the outset. Thus, we set up the
computational environment with {rix} first, then create and run the scripts, and
manuscript inside it. Last, we show how another researcher could reproduce it.

## Step 1: Installing Nix and {rix}

Both Nix and the {rix} R package need to be installed. The quick start below
covers the common case, but the complete, up-to-date instructions are maintained
in the {rix} documentation: Linux and Windows
(<https://docs.ropensci.org/rix/articles/setting-up-linux-windows.html>) and
macOS (<https://docs.ropensci.org/rix/articles/setting-up-macos.html>).

The installation is the same across operating systems, with one prerequisite on
Windows: Nix runs inside the Windows Subsystem for Linux 2 (WSL2), which must be
installed and configured first (see the {rix} documentation above). On all
systems, install Nix by running the Determinate Systems installer in a
terminal[^5]. To open it, on macOS, go to Applications → Utilities → Terminal.
For Windows, launch your WSL2 (Ubuntu) shell from the Start menu. For Linux,
open your distribution's terminal application. Then run (@lst-install-nix):

[^5]: A terminal is a text-based way to give your computer instructions. Readers
    new to the command line may find the following link a useful primer:
    <https://swcarpentry.github.io/shell-novice/>, as suggested in
    @wiebels_moreau_2021

::: {#lst-install-nix}
\small

``` bash
curl --proto '=https' --tlsv1.2 -sSf \
  -L https://install.determinate.systems/nix | \
  sh -s -- install
```

Installing Nix \normalsize
:::

Once Nix is installed[^6], there are two ways to access {rix}, depending on
whether R is already installed on your system. In this tutorial, we proceed as
if R was already installed[^7], as that most likely will be the case for most
readers (@lst-install-rix-cran):

[^6]: It is worth noting that {rix} can generate Nix expressions even without
    Nix installed on your system: you can write a `default.nix` file without
    Nix, but you cannot build or enter the resulting environment unless Nix is
    installed [@rodrigues_baumann_2025].

[^7]: We, however, recommend uninstalling your local R and letting Nix manage R,
    R packages, and other tools entirely. This approach avoids potential
    conflicts between system-installed and Nix-managed software, an issue we
    will illustrate later in this tutorial. See the {rix} documentation for more
    details.

```{r}
#| lst-label: lst-install-rix-cran
#| lst-cap: "Installing {rix} From CRAN or Development Version"
#| eval: false
#| echo: true
# CRAN version
install.packages("rix")
# Development version
install.packages(
  "rix",
  repos = c(
    "https://ropensci.r-universe.dev"
  )
)
```

## Step 2: Specifying the Computational Environment With `rix()` {#sec-specifying-environment}

We first begin by creating a folder for this project (here named
`Why-risk-it-when-you-can-rix-it`) that contains only two empty subfolders,
`Simulation` and `Manuscript`. We have not written any simulation scripts or the
manuscript yet. The only file we write by hand for now is `gen-env.R`, which
uses the `rix()` function from the {rix} package to produce a `default.nix`
file, a declarative specification of all software dependencies the project
needs.

We establish the full environment the simulation needs up front: the R version,
an IDE, and the R packages the scripts will use. You can always add more later,
but doing so means editing `gen-env.R` and following the steps we are about to
illustrate. The simulation specification is shown in @lst-rix-env-sim (the
complete version, which also covers the manuscript, is in the project
repository, <https://doi.org/10.5281/zenodo.21098114>, as `gen-env.R`):

\small

```{r}
#| lst-label: lst-rix-env-sim
#| lst-cap: "Simulation Environment: R, RStudio, and the R Packages"
#| eval: false
#| echo: true

library(rix)

# choose the path to your project
# here, the "Why-risk-it-when-you-can-rix-it" folder (root)
project_directory <- "."

rix(
  date = "2026-01-14", # recommendation: pin a date
  r_pkgs = c(
    # all R packages the simulation scripts need
    "rix",
    "marginaleffects",
    "simhelpers",
    "rvinecopulib",
    "doParallel",
    "doRNG",
    "ggplot2",
    "cowplot",
    "dplyr"
  ),
  ide = "rstudio", # editor to include
  project_path = project_directory,
  overwrite = TRUE
)
```

\normalsize

### Environment Specification

The `rix()` function constructs the specification through a series of
parameters that collectively describe the computational environment. Each
parameter serves a distinct purpose in defining the environment's
characteristics, and we will clarify those throughout this tutorial section. For
a broader overview of the function, see the {rix} documentation.

#### Specifying the R Version

Researchers must first determine which version of R to use. This can be
accomplished in two ways: The `r_ver` argument accepts an exact version string
(e.g., "4.3.3") or special designations such as "latest-upstream" for the most
recent stable release. Alternatively, the `date` argument specifies a particular
date (e.g., "2024-11-15"), which ensures that R and all packages correspond to
the versions available on that date. The date-based approach is generally
preferable for reproducibility, as it captures a complete snapshot of the R
ecosystem at a single point in time. For this tutorial, as shown on top, we use
the `date` parameter to ensure temporal consistency across all software
components [@rodrigues_baumann_2025].

\small

```{r}
#| lst-label: lst-date-example
#| lst-cap: "Example for the Date Argument"
#| eval: false
#| echo: true

date = "2026-01-14"
```

\normalsize

#### Declaring R Package Dependencies

The `r_pkgs` argument accepts a character vector listing all required R packages
by their CRAN names. These packages will be installed from the version
repository corresponding to the specified date or R version. It is important to
list all packages that the analysis will load directly; dependencies of these
packages are automatically resolved by Nix.

\small

```{r}
#| lst-label: lst-r-pkgs-example
#| lst-cap: "Example for the r_pkgs Argument"
#| eval: false
#| echo: true

r_pkgs = c(
  "rix", "marginaleffects", "simhelpers", "rvinecopulib",
  "doParallel", "doRNG", "ggplot2", "cowplot", "dplyr"
)
```

\normalsize

For packages requiring specific
versions not corresponding to the chosen date, researchers can specify exact
versions using the syntax `"packagename@version"` (e.g., `"ggplot2@2.2.1"`). For
packages available only on GitHub or other Git repositories, the `git_pkgs`
argument accepts a list structure containing repository URLs and specific commit
hashes. For example:

\small

```{r}
#| lst-label: lst-git-pkgs-example
#| lst-cap: "Example for the git_pkgs Argument"
#| eval: false
#| echo: true

git_pkgs = list(
  package_name = "marginaleffects",
  repo_url = "https://github.com/vincentarelbundock/marginaleffects",
  commit = "304bff91dc31ae28b227a8485bfa4f7bdc86d625"
)
```

\normalsize

This ensures that exact development versions are obtained
[@rodrigues_baumann_2025]. For our simulation study, all packages were used with
their CRAN versions. Appendix B gives the reasons for adding each package in
`r_pkgs`.

#### Configuring the Development Environment

The `ide` parameter controls whether an integrated development environment (IDE)
is included in the Nix environment, allowing users to interactively develop and
run code within their editor of choice. When `ide` is specified, the project can
be opened directly in the corresponding IDE, with all dependencies provided by
the Nix environment. Most readers likely already have an IDE installed locally.
Here we instead show how to open RStudio from the rix specification, which does
not require RStudio to be installed on your system. For example, setting
`ide = "rstudio"` installs a project-specific version of RStudio inside the Nix
environment. This is required for RStudio because, unlike most other editors, it
cannot attach to an external Nix shell unless it is itself installed via Nix. On
macOS, RStudio is only available through Nix for R versions 4.4.3 or later (or
environments dated 2025-02-28 or later). For earlier R versions, alternative
editors must be used. Other supported IDEs include Positron
(`ide = "positron"`), Visual Studio Code (`ide = "code"`), and command-line
interfaces such as Radian (`ide = "radian"`). These tools are installed directly
within the Nix environment using the `ide` parameter in this tutorial[^8]. 
All IDEs installed via Nix are project-specific and do not interfere with 
system-wide installations.

[^8]: Alternatively, users may rely on an editor already installed on their
    system by setting `ide = "none"` and using `direnv`, a
    lightweight utility that automatically activates the project's Nix
    environment, and thus its R version and packages, when navigating into the
    project directory (via a `.envrc` file), so an existing editor runs against
    it. As explained previously, in the specific case of RStudio, it must be installed
    using Nix.

\small

```{r}
#| lst-label: lst-ide-example
#| lst-cap: "Example for the ide Argument"
#| eval: false
#| echo: true

ide = "rstudio"
```

\normalsize

#### Setting File Output Parameters

The `project_path` parameter indicates where the `default.nix` file should be
written ("." denotes the current directory), while `overwrite` controls whether
an existing file should be replaced. Adding to this, setting `print = TRUE`,
which is another argument, displays the generated specification in the console
for immediate verification [@rodrigues_baumann_2025].

\small

```{r}
#| lst-label: lst-file-output-example
#| lst-cap: "Example for the project_path and Overwrite Arguments"
#| eval: false
#| echo: true

project_path = ".",
overwrite = TRUE
```

\normalsize

## Step 3: Generating the Environment Specification

After defining the computational environment, the `rix()` function must be
executed to generate the `default.nix` file. This can be done interactively by
running `rix()` (@lst-rix-env-sim). The file is written to the folder given in
`project_path`, so you can confirm it succeeded by checking that this folder now
contains `default.nix`.

## Step 4: Building and Using the Reproducible Environment

Once Step 3 is complete, we build the reproducible environment by navigating to
the project directory in the terminal. In our case, this is the
`Why-risk-it-when-you-can-rix-it` folder. From there, run the following command
(@lst-nix-build):

::: {#lst-nix-build}
\small

``` bash
# in the terminal, from the project directory:
nix-build
```

Building the Nix environment \normalsize
:::

The expected output should look similar to (@lst-nix-build-output):

::: {#lst-nix-build-output}
\small

``` bash
unpacking 'https://github.com/rstats-on-nix/nixpkgs/archive/2025-08-25.tar.gz'
  into the Git cache...
warning: ignoring untrusted substituter...
warning: ignoring the client-specified setting...
/nix/store/qa7fq20m2f94szsnqzciwv8h4n81w43v-nix-shell
```

Expected output from `nix-build` \normalsize
:::

This command builds the environment according to the specification. The first
execution will download and install all required packages, which may take a few
minutes depending on network speed and system resources. Subsequent builds use
cached packages and complete in seconds. Upon successful completion, a path to
the constructed environment in the Nix store is printed (here,
`/nix/store/qa7fq20m2f94szsnqzciwv8h4n81w43v-nix-shell`), and a symbolic link
named `result` appears in the project directory pointing to this location. In
our illustration, this build already includes the R packages declared for the
simulation (@lst-rix-env-sim).

Note that the warnings indicate that you are not configured as a trusted user,
so Nix cannot use the rstats-on-nix binary cache and will instead compile
packages from source, which is slower. To enable binary caching, see the {rix}
documentation: <https://docs.ropensci.org/rix/articles/binary-cache.html>.

To activate the environment, run (@lst-nix-shell):

::: {#lst-nix-shell}
\small

``` bash
# in the terminal, from the project directory:
nix-shell --pure
```

Activating the Nix environment \normalsize
:::

The expected output (if you have configured yourself as a trusted user,
otherwise the same warnings will appear) should look similar to
(@lst-nix-shell-output):

::: {#lst-nix-shell-output}
``` bash
unpacking 'https://flakehub.com/f/DeterminateSystems/nixpkgs-weekly/...' 
  into the Git cache...
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$
```

Expected output from `nix-shell`
:::

This command drops the user into a shell where all specified packages and tools
are available. The `--pure` flag restricts the shell to what is declared in
`default.nix`; we return to this flag later in the tutorial. The shell prompt
changes to indicate that a Nix environment is
active (here, `[nix-shell:~/Why-risk-it-when-you-can-rix-it]$`). To verify that
R is being provided by Nix rather than a system installation, run `which R`.
This should return a path within `/nix/store/`. Moreover, from within the Nix
shell, users can launch their IDE by typing its name (e.g., `rstudio` or
`positron`), which opens the IDE with the Nix environment active
(@lst-shell-rs):

::: {#lst-shell-rs}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$ rstudio
```

Activating RStudio \normalsize
:::

Note that launching RStudio this way does not automatically open the specific
project you are working in. We therefore recommend creating an RStudio project
file (`.Rproj`) and opening that file when using Nix, so that RStudio is
correctly associated with the intended project and environment.

## Step 5: Developing and Making the Simulation Reproducible

With RStudio open from the Nix environment, we next write the five simulation 
`.R` scripts inside the Simulation subfolder. Note that one does not need to use 
RStudio, but we illustrate it as most readers probably make use of an IDE.
Because we declared the simulation packages up front
(@lst-rix-env-sim), they are already available, and the scripts run against 
exactly those versions. Researchers are rarely aware of every package a project 
will need in advance, but, as highlighted before, packages can be added 
"on-the-go": as the code takes shape and calls for another package, we add it 
to `r_pkgs` in `gen-env.R` and rerun `rix()`. It is crucial that, after 
editing `gen-env.R`, we
rebuild with `nix-build` and re-enter with `nix-shell --pure` so the new
packages become available.

### Running the Simulation and Results

With the environment built, we run the simulation from inside the Nix shell so
that it uses exactly the R version and packages declared in `default.nix`. This
is how the data generation, results, and plots reported in this manuscript 
were produced. For example, `03_run_simulation.R` begins by loading the 
required packages and sourcing the other scripts it needs:

```{r}
#| lst-label: lst-run-simulation-code
#| lst-cap: "Code for Running the Simulation"
#| eval: false
#| echo: true

library(marginaleffects)
...
# Source helper functions
source("Simulation/01_data_generation.R")
...
```

The simulation needs nothing beyond the environment built above
(@lst-rix-env-sim). From within the Nix shell, the simulation runs
(@lst-run-simulation) as follows:

::: {#lst-run-simulation}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$
  Rscript Simulation/03_run_simulation.R
```

Running the complete simulation workflow \normalsize
:::

In the same way, we run `04_performance_metrics.R` (@lst-run-performance), which
loads the simulation results (in `sim_results.rds`) and calculates the
performance metrics:

::: {#lst-run-performance}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$
  Rscript Simulation/04_performance_metrics.R
```

Running the performance metrics calculation \normalsize
:::

Similarly, `05_plots.R` (@lst-run-plots) uses those saved metrics (in
`performance_summary.rds`) to create the plots:

::: {#lst-run-plots}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$
  Rscript Simulation/05_plots.R
```

Generating the visualization plots \normalsize
:::

Because these results and plots come entirely from running the scripts inside
the Nix environment, they are reproducible. This approach does, however, rely on
running the scripts in sequence by hand. It therefore guarantees a reproducible
environment but does not formalize the analytical pipeline itself, leaving the
dependencies between scripts implicit in the code rather than explicitly
declared.

## Step 6: Developing and Making the Complete Manuscript Reproducible

To also build the manuscript, we extend the simulation environment
(@lst-rix-env-sim) with the literate programming tools it lacks: Quarto,
{knitr}, and a set of LaTeX packages. We add these arguments to `gen-env.R` and
rebuild. @lst-rix-env-manuscript shows only what is added:

\small

```{r}
#| lst-label: lst-rix-env-manuscript
#| lst-cap: "Manuscript Tools Added to the Simulation Environment"
#| eval: false
#| echo: true

rix(
  # ...all simulation arguments from @lst-rix-env-sim, plus:
  r_pkgs = c(..., "quarto", "knitr", "svglite"), # add to existing vector
  system_pkgs = c("quarto"), # command-line Quarto
  tex_pkgs = c(
    "amsmath",
    "ninecolors",
    "apa7",
    "scalerel",
    "threeparttable",
    "threeparttablex",
    "endfloat",
    "environ",
    "multirow",
    "tcolorbox",
    "pdfcol",
    "tikzfill",
    "fontawesome5",
    "framed",
    "newtx",
    "fontaxes",
    "xstring",
    "wrapfig",
    "tabularray",
    "siunitx",
    "fvextra",
    "geometry",
    "setspace",
    "fancyvrb",
    "anyfontsize"
  )
)
```

\normalsize

#### Including System-level Dependencies

Nix handles non-R software in two ways. The system-level libraries your R
packages depend on, such as BLAS/LAPACK, are resolved automatically and need not
be listed. The {svglite} package, for example, depends on system libraries
(libpng and more), which we never list. The second way is the `system_pkgs`
parameter, which lists software that is not an R package, typically command-line
tools you run yourself inside the environment, such as Git or Quarto. In our
project this is only Quarto. Any software available through Nix can be requested
this way (searchable at <https://search.nixos.org/packages>), and, as with R
packages, a missing tool can simply be added to `system_pkgs` and the
environment rebuilt. Note the `apaquarto` extension we use for APA formatting is
a separate case: it is a Quarto extension stored in the project's `_extensions/`
directory rather than a Nix package [@rodrigues_baumann_2025].

#### Specifying LaTeX Packages

The `tex_pkgs` parameter lists the LaTeX packages needed for PDF compilation.
When any are listed, Nix includes a minimal TeXLive distribution
(`scheme-small`) and adds the requested packages on top. Finding the exact set
may take some trial and error, but Quarto's error messages during rendering
help by indicating which packages are missing. Appendix B gives also the 
reasons for adding each package in `tex_pkgs`.

After rebuilding with `nix-build` and re-entering with `nix-shell --pure`,
Quarto is
available. We write the manuscript (`article.qmd`) in a `Manuscript` subfolder,
and render it from the terminal (@lst-quarto-render):

::: {#lst-quarto-render}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$ 
  quarto render Manuscript/article.qmd
```

Rendering the manuscript with Quarto \normalsize
:::

This command executes all code chunks in the manuscript (i.e., loads the R
packages, runs the data generation and estimation), calculates and incorporates
results (e.g., @tbl-results) and figures (e.g., Figure 1), and generates a
formatted manuscript following APA style guidelines via the `apaquarto` extension
[@schneider_2024]. This extension is saved in the project directory already. To
add it to a project of your own, run the following inside the Nix shell, where
Quarto is provided by the environment and therefore needs no separate
installation (@lst-apaquarto-install):

::: {#lst-apaquarto-install}
\small

``` bash
[nix-shell:~/Why-risk-it-when-you-can-rix-it]$
  quarto add wjschne/apaquarto
```

Adding the apaquarto extension \normalsize
:::

The final document (which can be saved as a .docx, .pdf, or .html) is saved
directly in the project subfolder[^9]. Because Quarto is installed as a
command-line tool in our Nix specification, the rendering occurs entirely
within a fully reproducible environment, ensuring consistent output across
machines. If desired, the manuscript
can also be reproduced interactively by opening the project folder in the user’s
preferred IDE.

[^9]: Although we use `apaquarto` in this example, many alternative manuscript
    templates are available, and Nix is agnostic to the specific template
    employed, provided the necessary extensions are installed.

As highlighted before, we included the simulation code as code chunks for this
manuscript. This was done simply for illustration purposes. In a real-world
simulation study, users could for example run the simulation as in the previous
sections and then load the needed results (produced within the Nix shell) in the
manuscript. However, the same remark holds with respect to the formalized
workflow.

At this point, it is also worth noting that Nix shells do not fully isolate you
from your existing system by default. For R users, this has a practical
implication: packages installed in your regular R library (outside of Nix) could
potentially be loaded when running R from within the Nix environment. The {rix}
package addresses this automatically. When you call `rix()`, it also executes
`rix_init()`, which creates a project-specific `.Rprofile`. This file configures
R to ignore external package libraries and also disables `install.packages()`
within the environment. The rationale is straightforward: any new packages
should be added to `default.nix` and the environment rebuilt, preserving full
reproducibility [@rodrigues_baumann_2025]. In addition, for stricter
isolation[^10] that also prevents access to other system programs not specified
in `default.nix`, in this tutorial, we activated the environment with the 
`--pure` flag from the start (@lst-nix-shell). A related caveat specific to 
macOS is discussed in the Discussion section below.

[^10]: For example, when preparing this manuscript without the `--pure` flag,
    `quarto render` worked successfully. However, when using the `--pure` flag,
    the build failed. Running `quarto check` from within the Nix shell (i.e.,
    `nix-shell --run "quarto check"`) revealed that Quarto was still accessing
    the system's LaTeX installation (`/Library/TeX/texbin`) rather than being
    restricted to only what was specified in `default.nix`.

## Reproducing the Simulation Study and Manuscript

Having built the project as its authors, we now switch to the perspective of a
reader who wants to reproduce it. Because the environment is bundled with the
project, reproduction requires almost nothing beyond the project folder itself.
A reader can download the entire project from its repository
(<https://doi.org/10.5281/zenodo.21098114>). The folder already contains
`default.nix`, thus they do not need the authoring steps (Steps 1 to 3). With
Nix installed, they need only build and enter the environment as in Step 4
(`nix-build`, then `nix-shell --pure`), which guarantees that all dependencies
(i.e., the R version, packages, and system dependencies) match those specified 
in `default.nix`. From within that environment, reproduction proceeds on two 
levels. The simulation can be reproduced by running its scripts in order, 
regenerating the results and figures. The complete manuscript 
(its text together with the embedded results, tables, and figures) can be 
reproduced in the same environment as in @lst-quarto-render, which re-executes 
every code chunk and regenerates the formatted manuscript end to end.

```{r}
#| lst-label: lst-data-generation
#| lst-cap: "Data Generation Function"
#| echo: false
#| message: false

# fixed data-generating process (DGP) parameters
# parameters for treatment assignment model: X1 = alpha0 + alpha1*X2 + alpha2*X2^2 + error
alpha0 <- 0 # intercept for treatment equation
alpha1 <- 0.5 # linear effect of confounder X2 on treatment X1
alpha2 <- 0.2 # quadratic effect of confounder X2 on treatment X1

# parameters for outcome model: logit(P(Y=1)) = beta0 + beta1*X1 + gamma1*X2 + gamma2*X2^2
beta0 <- -0.5 # intercept for outcome equation
beta1 <- 0.7 # causal effect of treatment X1 on outcome Y (parameter of interest)
gamma1 <- -0.4 # linear effect of confounder X2 on outcome Y

# function to generate one dataset
# arguments:
#   n: sample size
#   gamma2: quadratic effect of confounder on outcome (varies by condition)
gen_data <- function(n, gamma2) {
  # generate independent uniform pairs using {rvinecopulib} (C++ backend)
  # this is equivalent to rnorm() but demonstrates system-level dependencies
  u <- rbicop(n, "indep", 0, numeric(0))

  # transform uniform marginals to standard normal via inverse CDF
  X2 <- qnorm(u[, 1]) # confounder
  error_X1 <- qnorm(u[, 2]) # error term for treatment equation

  # generate treatment variable (confounded by X2)
  X1 <- alpha0 + alpha1 * X2 + alpha2 * X2^2 + error_X1

  # compute true linear predictor for outcome model
  eta_true <- beta0 + beta1 * X1 + gamma1 * X2 + gamma2 * X2^2

  # convert to probability via logistic function
  p <- plogis(eta_true)

  # generate binary outcome
  y <- rbinom(n, size = 1, prob = p)

  # dataset with outcome and predictors
  data.frame(y = y, x1 = X1, x2 = X2)
}
```

```{r}
#| lst-label: lst-model-estimation
#| lst-cap: "Model Estimation Functions"
#| echo: false
#| message: false

# function to run one simulation replicate using the MISSPECIFIED model
# arguments:
#   n: sample size
#   gamma2: quadratic effect of confounder on outcome
run_one_rep <- function(n, gamma2) {
  # generate dataset for this replicate
  dat <- gen_data(n, gamma2)

  # fit misspecified model (omits X2^2 term, causing residual confounding)
  fit_misspec <- glm(y ~ x1 + x2, data = dat, family = binomial())

  # compute average causal effect (ACE) using {marginaleffects}
  ace_est <- avg_slopes(fit_misspec, variables = "x1")

  # point estimate, standard error, and 95% CI bounds
  data.frame(
    est = ace_est$estimate[1],
    se = ace_est$std.error[1],
    lo = ace_est$conf.low[1],
    hi = ace_est$conf.high[1]
  )
}

# function to compute the "true" ACE using the correctly specified model
# uses a large sample to approximate the population quantity
# arguments:
#   gamma2_val: quadratic effect of confounder on outcome
#   seed: random seed for reproducibility
compute_true_ace <- function(gamma2_val, seed = 12345) {
  set.seed(seed)

  # generate large dataset (n = 200,000) to approximate population
  bigdat <- gen_data(2e5, gamma2_val)

  # fit correctly specified model (includes X2^2 term)
  fit_correct <- glm(y ~ x1 + x2 + I(x2^2), data = bigdat, family = binomial())

  # true ACE estimate
  avg_slopes(fit_correct, variables = "x1")$estimate[1]
}
```

```{r}
#| lst-label: lst-main-simulation-loop
#| lst-cap: "Main Simulation"
#| echo: false
#| message: false
#| output: false

# simulation parameters
nsim <- 100 # replications per condition
ncore <- 1 # CPU cores for parallel processing
master_seed <- 12345 # seed

# simulation factors
n_values <- c(50, 100, 2000) # sample sizes to evaluate
gamma2_values <- c(0, 0.3, 0.8) # confounding severity levels
gamma2_labels <- c("none", "mild", "severe") # labels for gamma2 levels

# full factorial design (3 x 3 = 9 conditions)
designs <- expand.grid(
  n = n_values,
  gamma2 = gamma2_values,
  stringsAsFactors = FALSE
)

# descriptive labels for confounding severity
designs$confound_label <- factor(
  designs$gamma2,
  levels = gamma2_values,
  labels = gamma2_labels
)

# compute true ACE values for each gamma2 condition
# these serve as the reference for evaluating estimator performance
cat("Computing true ACE values\n")
true_ACE_values <- sapply(gamma2_values, compute_true_ace)
names(true_ACE_values) <- as.character(gamma2_values)
print(true_ACE_values)

# parallel processing backend
cl <- makeCluster(ncore)
registerDoParallel(cl)

# fix the master seed so the {doRNG} streams below are reproducible and
# independent of the number of cores
set.seed(master_seed)

# run simulation across all conditions
all_results <- vector("list", nrow(designs))

# per-iteration RNG stream states, so any single repetition within a
# condition can later be reproduced in isolation (see attr(res_list, "rng"))
rng_states <- vector("list", nrow(designs))

for (k in seq_len(nrow(designs))) {
  # extract current condition parameters
  n_k <- designs$n[k]
  gamma2_k <- designs$gamma2[k]
  label_k <- as.character(designs$confound_label[k])

  cat("\nCondition:", label_k, "(gamma2 =", gamma2_k, "), n =", n_k, "\n")

  # true ACE for this condition
  true_ACE_k <- true_ACE_values[as.character(gamma2_k)]

  # run nsim replications in parallel with reproducible RNG via {doRNG}
  res_list <- foreach(
    i = seq_len(nsim),
    .packages = c("marginaleffects", "rvinecopulib")
  ) %dorng%
    {
      run_one_rep(n_k, gamma2_k)
    }

  # store the stream state {doRNG} assigned to each repetition
  rng_states[[k]] <- attr(res_list, "rng")

  # combine replications into data frame
  res <- do.call(rbind, res_list)

  # add condition metadata to results
  res$true_param <- true_ACE_k
  res$n <- n_k
  res$gamma2 <- gamma2_k
  res$confound_label <- label_k

  all_results[[k]] <- res
}

# clean up parallel backend
stopCluster(cl)

# all conditions into single data frame
sim_results <- do.call(rbind, all_results)
```

```{r}
#| lst-label: lst-performance-metrics
#| lst-cap: "Performance Metrics"
#| echo: false
#| message: false

# unique simulation conditions
conditions <- unique(sim_results[, c("n", "gamma2", "confound_label")])

# initialize list to store performance metrics for each condition
summary_list <- vector("list", nrow(conditions))

# loop over each condition to compute performance metrics
for (k in seq_len(nrow(conditions))) {
  # subset results for current condition
  res <- subset(
    sim_results,
    n == conditions$n[k] &
      gamma2 == conditions$gamma2[k]
  )

  # absolute performance metrics using {simhelpers}
  # includes bias, variance, MSE, and RMSE (with MCSEs)
  perf_abs <- calc_absolute(
    data = res,
    estimates = est,
    true_param = true_param,
    criteria = c("bias", "variance", "mse", "rmse")
  )

  # relative performance metrics
  # includes relative bias and relative RMSE (with MCSEs)
  perf_rel <- calc_relative(
    data = res,
    estimates = est,
    true_param = true_param,
    criteria = c("relative bias", "relative rmse")
  )

  # coverage and CI width metrics
  # evaluates 95% CI performance (with MCSEs)
  perf_cov <- calc_coverage(
    data = res,
    lower_bound = lo,
    upper_bound = hi,
    true_param = true_param,
    criteria = c("coverage", "width")
  )

  # combine condition info with all performance metrics
  summary_list[[k]] <- cbind(
    conditions[k, ],
    true_ace = unique(res$true_param),
    perf_abs,
    perf_rel,
    perf_cov
  )
}

# all conditions into final summary data frame
performance_summary <- do.call(rbind, summary_list)
```

```{r}
#| lst-label: lst-fig-performance
#| lst-cap: "Code for generating the performance metrics visualization"
#| label: fig-performance
#| fig-cap: "Performance of the average causal effect (ACE) estimator across
#| sample sizes and confounding severity. Panel A shows relative bias, Panel B
#| shows relative root mean squared error (RMSE), Panel C shows coverage
#| probability of 95% confidence intervals (CI; dashed line at nominal 0.95
#| level), and Panel D shows average CI width. Results demonstrate that model
#| misspecification induces systematic bias that persists across sample sizes,
#| while increasing sample size improves precision but not accuracy under
#| misspecification."
#| fig-width: 10
#| fig-height: 8
#| echo: false
#| warning: false

# panel A: absolute bias plot (not used in final figure)
# shows raw difference between estimated and true ACE
p_bias <- ggplot(
  performance_summary,
  aes(x = n, y = bias, color = confound_label, group = confound_label)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "Absolute Bias",
    color = "Confounding\nnon-linearity",
    title = "Absolute Bias from Omitting X²"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# panel A (final): relative bias plot
# ratio of estimated to true ACE (1 = unbiased)
p_rel_bias <- ggplot(
  performance_summary,
  aes(x = n, y = rel_bias, color = confound_label, group = confound_label)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "Relative Bias",
    color = "Confounding\nnon-linearity",
    title = "Relative Bias from Omitting X²"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# absolute RMSE plot (not used in final figure)
# raw root mean squared error
p_rmse <- ggplot(
  performance_summary,
  aes(x = n, y = rmse, color = confound_label, group = confound_label)
) +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "RMSE of ACE estimator",
    color = "Confounding\nnon-linearity",
    title = "Absolute RMSE of ACE Estimation"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# panel B: relative RMSE plot
# RMSE scaled by true parameter value
p_rel_rmse <- ggplot(
  performance_summary,
  aes(x = n, y = rel_rmse, color = confound_label, group = confound_label)
) +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "Relative RMSE",
    color = "Confounding\nnon-linearity",
    title = "Relative Accuracy of ACE Estimation"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# panel C: coverage probability plot
# proportion of 95% CIs containing the true ACE
# dashed line at 0.95 = nominal coverage level
p_coverage <- ggplot(
  performance_summary,
  aes(x = n, y = coverage, color = confound_label, group = confound_label)
) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "gray50") +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "Coverage (95% CI)",
    color = "Confounding\nnon-linearity",
    title = "Coverage Under Residual Confounding"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_minimal() +
  theme(legend.position = "right")

# panel D: confidence interval width plot
# average width of 95% CIs (measures precision)
p_width <- ggplot(
  performance_summary,
  aes(x = n, y = width, color = confound_label, group = confound_label)
) +
  geom_point(size = 3) +
  geom_line() +
  labs(
    x = "Sample size (n)",
    y = "Average CI width",
    color = "Confounding\nnon-linearity",
    title = "Precision of CI"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# combine plots into 2x2 grid with shared legend using {cowplot}
# extract legend from one plot to use as shared legend
legend <- get_legend(
  p_rel_bias + theme(legend.position = "right")
)

# arrange panels and legend side by side
plot_grid(
  plot_grid(p_rel_bias, p_rel_rmse, p_coverage, p_width, ncol = 2, nrow = 2),
  legend,
  rel_widths = c(3, 0.4)
)
```

```{r}
#| lst-label: lst-tbl-results
#| lst-cap: "Code for generating the performance metrics table"
#| label: tbl-results
#| tbl-cap: "Performance metrics for the average causal effect (ACE) estimator
#| across simulation conditions. Values in parentheses are Monte Carlo standard
#| errors (MCSE). RMSE = root mean squared error; CI = confidence interval."
#| echo: false

# prepare data for display table
# select relevant columns including estimates and their MCSEs
results_display <- performance_summary %>%
  select(
    n,
    confound_label,
    rel_bias,
    rel_bias_mcse,
    rel_rmse,
    rel_rmse_mcse,
    coverage,
    coverage_mcse,
    width,
    width_mcse
  ) %>%
  mutate(
    # rename columns for display
    `Sample Size` = n,
    `Confounding` = confound_label,
    # format each metric as "estimate (MCSE)" with 3 decimal places
    `Relative Bias` = sprintf("%.3f (%.3f)", rel_bias, rel_bias_mcse),
    `Relative RMSE` = sprintf("%.3f (%.3f)", rel_rmse, rel_rmse_mcse),
    `Coverage` = sprintf("%.3f (%.3f)", coverage, coverage_mcse),
    `CI Width` = sprintf("%.3f (%.3f)", width, width_mcse)
  ) %>%
  # only the formatted columns for final display
  select(
    `Sample Size`,
    `Confounding`,
    `Relative Bias`,
    `Relative RMSE`,
    `Coverage`,
    `CI Width`
  )

# render table
kable(results_display, align = "lccccc", row.names = FALSE)
```

# Additional Considerations for Simulation Studies

In this section, we turn to considerations that matter especially for simulation
studies but fall outside the scope of this tutorial, since they depend on the 
details of each study. The first concerns random draws and reproducibility. 
The second concerns the analytical pipeline, environment for multiple 
programming languages environment, and integration with tools researchers may 
already use. We do not cover these topics in depth. Our goal is to clarify how 
they relate to {rix} and to point readers toward resources for more advanced 
use cases.

### Reproducible Randomness and Parallel Execution for Data Generation

A simulation regenerates its data from a pseudo-random number generator. Its
reproducibility therefore depends on controlling that generator. The standard
advice is to set the seed once and store it, which ensures that the same code
reproduces the same sequence of draws [@morris_etall_2019]. Simulation studies
are typically organized as a set of conditions, each run for many repetitions,
and researchers often need to reproduce a single repetition within a given
condition. For this purpose, the state of the generator can additionally be
stored at the start of each repetition of each condition, rather than only the
initial seed. Any repetition can then be reproduced in isolation, which is
convenient for debugging [@morris_etall_2019]. In R, this state is stored in the
variable `.Random.seed`, and the procedure would look as follows
(@lst-rng-state):

```{r}
#| lst-label: lst-rng-state
#| lst-cap: "Storing and Restoring the Generator State"
#| eval: false
#| echo: true

# fix the starting position once
set.seed(123)

# 4 conditions, 1000 repetitions per condition
conditions <- expand.grid(n = c(50, 100), gamma2 = c(0, 0.3))
nsim <- 1000

# one slot per condition-repetition cell
states <- matrix(list(), nrow(conditions), nsim)
results <- matrix(NA_real_, nrow(conditions), nsim)

# nested loops shown for illustration
for (cond in seq_len(nrow(conditions))) {
  for (r in seq_len(nsim)) {
    # state at the start of cell (cond, r)
    states[[cond, r]] <- .Random.seed
    # data generation advances the state
    x <- rnorm(conditions$n[cond])
    results[cond, r] <- mean(x)
  }
}

# later, reproduce repetition 3 of condition 2 on its own
.Random.seed <- states[[2, 3]]
identical(mean(rnorm(conditions$n[2])), results[2, 3]) # TRUE
```

With a sufficient number of repetitions the particular seed does not affect the
conclusions. However, parallel processing complicates this [@siepe_etall_2024;
@morris_etall_2019]. In this case, a single seed may no longer be sufficient,
because each worker would advance its own copy of the generator and could draw
overlapping sequences. In practice, many researchers assign a distinct seed to
each condition, for example `set.seed(123 + i)` for condition `i`, and let the
repetitions within that condition draw sequentially from it. This keeps every
condition reproducible and might not be a problem for most studies (given the
properties of the default algorithm, i.e., Mersenne-Twister), but it does not by
itself *guarantee* that the random sequences of different conditions never
overlap. The recommended solution is to assign separate, non-overlapping random
number streams [@morris_etall_2019; @siepe_etall_2024]. R offers several
packages for this [@rcoreteam_parallel; @rstream; @doRNG], and we illustrate the
approach with {doRNG}, which implements streams for {foreach} [@foreach] through
the `%dorng%` operator (@lst-run-simulation-code). From a single master seed set
before the loop, {doRNG} derives one independent stream per iteration, bound to
the iteration index rather than to a worker. The number of cores (`ncore` in our
script) therefore changes only how iterations are scheduled, never which stream
each receives. Thus, a researcher who reruns the study on four cores reproduces
data generated on twenty cores. Because {doRNG} also records each stream
in the result's `rng` attribute, these states can be stored and later restored
to reproduce any single repetition in isolation, which the per-condition seed
alone cannot do (@lst-rng-streams). Researchers should store these states and,
preferably, make them available alongside the reported estimates.

```{r}
#| lst-label: lst-rng-streams
#| lst-cap: "Reproducible Parallel Streams with {doRNG}"
#| eval: false
#| echo: true

library(doRNG)

# makes the streams below reproducible
set.seed(123)

# one slot per condition
rng_states <- vector("list", nrow(designs))

# nested loop over conditions
for (k in seq_len(nrow(designs))) {
  res <- foreach(
    i = seq_len(nsim),
    .packages = c("marginaleffects", "rvinecopulib")
  ) %dorng%
    {
      run_one_rep(designs$n[k], designs$gamma2[k])
    }
  # per-repetition stream states for this condition
  rng_states[[k]] <- attr(res, "rng")
  all_results[[k]] <- do.call(rbind, res)
}

# later, reproduce repetition 10 of condition 3 on its own
RNGkind("L'Ecuyer-CMRG")
.Random.seed <- rng_states[[3]][[10]]
rep10 <- run_one_rep(designs$n[3], designs$gamma2[3])
```

### Analytical Pipelines: {targets} and {rixpress}

As mentioned, complex simulation studies often benefit from an analytical 
pipeline that tracks dependencies between computational steps, caches 
intermediate results, and enables selective re-execution when inputs change. 
Two complementary approaches exist within the Nix ecosystem.

#### Using {targets} Within Nix

The {targets} package [@landau_2021] provides analytical pipeline tooling for
R-based projects. To integrate {targets} with Nix, include `targets` in the
`r_pkgs` parameter of `rix()` and execute the pipeline within `nix-shell` using
`Rscript -e 'targets::tar_make()'`. A shell hook can also be added via the
`shell_hook` argument to run the pipeline automatically when entering the Nix
shell. This approach is ideal for projects that remain within R and do not
require different environments for different pipeline steps (see {rix}
documentation:
<https://docs.ropensci.org/rix/articles/reproducible-pipelines.html>).

#### Using {rixpress} for Multi-Language Pipelines

The {rixpress} package [@rixpress], a sister package to {rix}, uses Nix itself
as the build automation tool rather than operating within a Nix environment.
Each pipeline step becomes a Nix derivation, built in isolation with automatic
caching based on content. The key advantage emerges in multi-language workflows:
different steps can execute in different Nix-defined environments (e.g., one
step using a specific version of R, another using Python, another using Julia).
The interface, inspired by {targets}, uses functions like `rxp_r()`, `rxp_py()`,
and `rxp_jl()` to define pipeline steps (see {rixpress} documentation:
<https://docs.ropensci.org/rixpress/articles/intro-concepts.html>). The GitHub
repository for this article directs interested readers to a demonstration of
{rixpress} applied to this entire project.

### Multi-Language Environment Support

While this tutorial focuses on R, researchers working across multiple
programming languages can include Python or Julia in their environments. The
`py_conf` parameter accepts a list specifying a Python version and required
packages (@lst-python-pkgs-example). Similarly, `jl_conf` enables Julia package
installation. This capability is particularly useful, for example, for projects
requiring statistical computing in R alongside machine learning pipelines in
Python or numerical optimization in Julia [@rodrigues_baumann_2025] (see {rix}
documentation for more:
<https://docs.ropensci.org/rix/articles/installing-r-packages.html>).

```{r}
#| lst-label: lst-python-pkgs-example
#| lst-cap: "Including Python packages"
#| eval: false
#| echo: true

py_conf = list(py_version = "3.12", py_pkgs = c("polars", "pandas"))
```

### Converting Existing {renv} Projects

Researchers with existing {renv} projects can migrate using the `renv2nix()`
function, which reads an `renv.lock` file and generates an equivalent Nix
expression. This is particularly valuable for projects where {renv} encountered
system dependency issues or where stricter reproducibility guarantees are
desired. Unlike {renv}, which captures R package versions but not the R
interpreter or system-level dependencies, Nix manages all layers of the software
stack (see {rix} documentation:
<https://docs.ropensci.org/rix/articles/renv2nix.html>).

### Containerization with Docker

Nix and Docker are not necessarily mutually exclusive
[@rodrigues_baumann_2026_polyglot]. Researchers already using Docker do not need
to abandon it to benefit from Nix, as the two can be combined by using Nix
inside Docker containers to handle environment setup [@rodrigues_baumann_2025].
This is particularly useful for deployment to cloud platforms or
high-performance computing clusters where Nix may not be installed. In that
case, the environment is built with Nix but shipped inside a Docker image, so it
can be run with Docker alone (see {rix} documentation: 
<https://docs.ropensci.org/rix/articles/nix-inside-docker.html>).

# Discussion

Reproducibility in computational research is often treated as a matter of
transparency: making data and code available. This tutorial has argued that
transparency alone is insufficient without the ability to reliably reconstruct
the computational environments in which analyses are executed. For simulation
studies in particular, where results depend critically on software versions,
system-level dependencies, and random number generation, environment-level
reproducibility is not optional but essential.

By introducing Nix and the {rix} package, we demonstrated a practical and
accessible approach to fully specifying and rebuilding computational
environments for simulation-based research. This approach greatly increases the
chances that analyses and manuscripts can be reproduced across machines and over
time, moving reproducibility from an aspirational goal toward a verifiable
property of the research workflow.

Importantly, adopting environment reproducibility does not require abandoning
existing analytic practices. Nix is agnostic to programming language, editor,
workflow structure, and manuscript template, allowing researchers to retain
familiar tools while strengthening the reliability of their work. In this sense,
reproducible environments serve as enabling infrastructure, supporting rather
than replacing other best practices such as version control, analytical
pipelines, and transparent reporting.

Therefore, if reproducibility is to function as a cornerstone of cumulative 
science, then the ability to reconstruct computational environments must 
become a routine part of methodological practice. Nix leveraged by tools 
like {rix} lowers the barrier to achieving this goal. We hope this tutorial will 
help to normalize environment-level reproducibility as a standard component 
of rigorous computational research in psychological science and beyond.

## Limitations

The approach we describe has limitations worth noting, the most relevant of
which is specific to macOS. There, some parts of an environment still depend on
Apple's own system software, which lies outside the Nix store and is therefore
not pinned. As a consequence, an environment pinned to an older date can
occasionally stop building or loading after a macOS update. Because reproducing
an environment faithfully means keeping its pin unchanged, the reliable solution
is not to update the pin but to build and run the original environment through
Docker or on Linux (including WSL on Windows), where it no longer depends on the
host's macOS [@rodrigues_baumann_2026_polyglot]. This preserves exact
reproducibility while working around the platform's constraints (for
macOS-specific setup, see
<https://docs.ropensci.org/rix/articles/setting-up-macos.html>).

Moreover, a reproducible environment does not, on its own, make the workflow
itself transparent. This is particularly relevant for simulation studies, which
are often split across several files whose dependencies, and the order in which
they should run, are not always clear. Dedicated analytical pipeline tools can 
formalize these dependencies (see above), but at a minimum, clearly documenting 
how the files relate and in what order they run already helps others 
reproduce the study. Thus, we strongly advise that at least such a description is
provided.

Last, as highlighted, there may be components of simulation studies that
might be hard to reproduce by other researchers such as very computationally
intensive studies that demand a huge amount of computer power. The data
generation also might not yield identical numerical results due to the
interplay of multiple factors. However, making available the obtained results,
together with the pinned environment, would at least allow the reproducibility
of the analyses and manuscript, as illustrated in this article. This distinction
in itself is relevant because it avoids coarse statements about the 
reproducibility of simulation studies [@montoya_anderson_2026].

{{< pagebreak >}}

# References

::: {#refs}
:::

{{< include appendix_a.qmd >}}

{{< include appendix_b.qmd >}}
